---
title: "Problem Set 5"
author: "Tiffany Wu"
format: 
  html:
    code-fold: true
    code-summary: "Show the code"
    embed-resources: true
output: html_document
editor: source
mainfont: Times New Roman
---

## Github Repo Link to Problem Set 5:
[https://github.com/tiffany-wu/Stats506/tree/main/PS5](https://github.com/tiffany-wu/Stats506/tree/main/PS5) 

## Problem 1 - OOP Programming

Create a class to represent rational numbers. Do this using S4.

### a. For the rational class, define the following:

1. A constructor
2.A validator that ensures the denominator is non-zero.
3.A show method.
4.A simplify method, to obtain the simplest form (e.g. simplify(2/4) produces 1/2).
5.A quotient method (e.g. quotient(3/7) produces .42857143...). It should support a digits argument but only in the printing, not the returned result (Hint: what does print return?).
6.Addition, subtraction, multiplication, division. These should all return a rational.
7.You’ll (probably) need GCD and LCM as part of some of these calculations; include these functions using Rcpp. Even if you don’t need these functions for another calculation, include them.
```{r}
rm(list=ls()) 
```

```{r}
# Load Rcpp for GCD and LCM functions
library(Rcpp)

# Define GCD and LCM functions in C++
cppFunction("
#include <cmath>       // Include cmath for abs function
#include <numeric>     // Include numeric for std::gcd and std::lcm functions

int C_gcd(int x, int y) {
  return std::gcd(x, y);
}

int C_lcm(int x, int y) {
  return abs(x * y) / C_gcd(x, y);
}
")
```

```{r}
# Set up rational class

# OLD: Define GCD and LCM functions in C++ -- from class notes
#cppFunction("int C_gcd(int x, int y) {
#  return std::gcd(x, y);
#}
#int C_lcm(int x, int y) {
#  return abs(x * y) / C_gcd(x, y);}")

# Define the S4 class
setClass("Rational",
         slots = c(numerator = "numeric", denominator = "numeric"),
         prototype = list(numerator = 0, denominator = 1)
)

# Validity
setValidity("Rational", function(object) {
  if (object@denominator == 0) {
    return("Denominator cannot be zero.")
  }
  TRUE
})

# Simplify method-- this is just for internal use within constructor to obtain the simplest form
simplify_fraction <- function(numerator, denominator) {
  divisor <- gcd(numerator, denominator)
  numerator <- numerator / divisor
  denominator <- denominator / divisor
  list(numerator = numerator, denominator = denominator)
}

# Constructor
Rational <- function(numerator, denominator = 1) {
  if (denominator == 0) {
    stop("Error: Denominator cannot be zero.")
  }
  obj <- new("Rational", numerator = numerator, denominator = denominator)
  obj  # Return the valid object
}

# Show method
setMethod("show", "Rational", function(object) {
  cat(object@numerator, "/", object@denominator, "\n")
})

# Simplify method for external use
setGeneric("simplify", function(object) standardGeneric("simplify"))
setMethod("simplify", "Rational", function(object) {
  simplified <- simplify_fraction(object@numerator, object@denominator)
  new("Rational", numerator = simplified$numerator, denominator = simplified$denominator)
})

# Quotient
setGeneric("quotient", function(object, digits = NULL) standardGeneric("quotient"))
setMethod("quotient", "Rational", function(object, digits = NULL) {
  result <- object@numerator / object@denominator
  if (!is.null(digits) && is.numeric(digits)) {
    print(round(result, digits))
  } else {
    print(result)
  }
  invisible(result)
})

# Addition
setMethod("+", signature(e1 = "Rational", e2 = "Rational"), function(e1, e2) {
  lcm_den <- lcm(e1@denominator, e2@denominator)
  numerator <- e1@numerator * (lcm_den / e1@denominator) + e2@numerator * (lcm_den / e2@denominator)
  Rational(numerator, lcm_den)  # Automatically simplifies
})

# Subtraction
setMethod("-", signature(e1 = "Rational", e2 = "Rational"), function(e1, e2) {
  lcm_den <- lcm(e1@denominator, e2@denominator)
  numerator <- e1@numerator * (lcm_den / e1@denominator) - e2@numerator * (lcm_den / e2@denominator)
  Rational(numerator, lcm_den)  # Automatically simplifies
})

# Multiplication
setMethod("*", signature(e1 = "Rational", e2 = "Rational"), function(e1, e2) {
  Rational(e1@numerator * e2@numerator, e1@denominator * e2@denominator)  # Automatically simplifies
})

# Division
setMethod("/", signature(e1 = "Rational", e2 = "Rational"), function(e1, e2) {
  if (e2@numerator == 0) stop("Cannot divide by a rational number with zero numerator.")
  Rational(e1@numerator * e2@denominator, e1@denominator * e2@numerator)  # Automatically simplifies
})
```


### b. Use your rational class to create three objects:

r1: 24/6
 
r2: 7/230
 
r3: 0/4

```{r}
# Creating objects
r1 <- Rational(24, 6)    # This should simplify to 4/1
r2 <- Rational(7, 230)   # This will stay as 7/230
r3 <- Rational(0, 4)     # This should simplify to 0/1

r1 # 4/1
r2 # 7/230
r3 #0/1

```

 
Evaluate the following code (remember you can tell Quarto not to stop on errors):

```{r, error = TRUE}
r1
r3
r1 + r2
r1 - r2
r1 * r2
r1 / r2
r1 + r3
r1 * r3
r2 / r3
quotient(r1)
quotient(r2)
quotient(r2, digits = 3)
quotient(r2, digits = 3.14)
quotient(r2, digits = "avocado")
q2 <- quotient(r2, digits = 3)
q2
quotient(r3)
simplify(r1)
simplify(r2)
simplify(r3)
```


Show that your validator does not allow the creation of rational’s with 0 denominator, and check other malformed input to your constructor.
```{r, error = TRUE}
# Testing the validator for Rational objects with invalid inputs -- should all error
Rational(1, 0)
Rational("a", 2)
Rational(1, "b")
Rational("a", "b")
```


Note that there are a lot of choices to be made here. How are you going to store the class? Two numerics? A vector of length two? A formula? A string? What are users going to pass into the constructor? A string (“24/6”)? Two arguments? A vector?

There is no right answer to those questions. Make the best decision you can, and don’t be afraid to change it if your decision causes unforeseen difficulties.

You may not use any existing R functions or packages that would trivialize this assignment. (E.g. if you found an existing package that does this, or found a function that automatically produces the quotient or simplified version, that is not able to be used.)

Hint: It may be useful to define other functions that I don’t explicitly ask for.


##Problem 2 - plotly

Let’s revisit the art data from the last problem set. Use plotly for these.

```{r}
library(plotly)

# Load art sales df
art <- read.csv("df_for_ml_improved_new_market.csv")
```

### a. Regenerate your plot which addresses the second question from last time:

#### ii. Does the distribution of genre of sales across years appear to change?
You may copy your plot from last time, or copy my plot from the solutions, or come up with your own new plot.

```{r}
library(tidyverse)

# Get year and genre columns, clean up column names, and process two+ genres
genre <- art %>%
  dplyr::select(year, starts_with("Genre")) %>%
  rename_with(~ gsub("Genre___", "", .), starts_with("Genre___")) %>%
  mutate(two_plus = rowSums(select(., Photography, Print, Sculpture, Painting, Others), na.rm = TRUE)) %>%
  mutate("Two+ Genres" = ifelse(two_plus > 1, 1, 0)) %>%
  select(-two_plus)

# Reshape to long format
genre_long <- genre %>%
  pivot_longer(
    cols = -year,  
    names_to = "genre",  
    values_to = "count"
  )

# Summarize total sales for each genre per year
genre_summary <- genre_long %>%
  group_by(year, genre) %>%
  summarize(total_sales = sum(count, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(genre = factor(genre, levels = c("Photography", "Print", "Sculpture", "Painting", "Others", "Two+ Genres")))

# Interactive stacked bar chart using Plotly
plot_genre_distribution <- plot_ly(
  data = genre_summary,
  x = ~year,
  y = ~total_sales,
  color = ~genre,
  type = "bar",
  colors = "Set3"
) %>%
  layout(
    title = "Distribution of Genre of Sales Across Years",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Total Sales"),
    barmode = "stack"
  )

plot_genre_distribution
```


### b. Generate an interactive plot with plotly that can address both of these questions from last time:

### i. Is there a change in the sales price in USD over time?

### ii. How does the genre affect the change in sales price over time?

This should be a single interactive plot, with which a user can manipulate the view to be able to look at change over time overall, or by genre.

```{r}
# Calculate average sales price per genre per year
price_trend <- art %>%
  dplyr::select(year, price_usd, starts_with("Genre")) %>%
  # Get rid of the Genre___ start so we can just use the genre name for plot
  rename_with(~ gsub("Genre___", "", .), starts_with("Genre___")) %>%
  mutate(genre = case_when(Photography == 1 ~ "Photography",
                           Print == 1 ~ "Print",
                           Sculpture == 1 ~ "Sculpture",
                           Painting == 1 ~ "Painting",
                           Others == 1 ~ "Other")) %>%
  group_by(year, genre) %>%
  summarize(
    mean_price = mean(price_usd, na.rm = TRUE),
    median_price = median(price_usd, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  mutate(genre = factor(genre, levels = c("Photography", "Print", "Sculpture", "Painting", "Other")))


# Create an interactive plotly plot with both mean and median prices
plot <- plot_ly(data = price_trend) %>%
  add_trace(
    x = ~year, y = ~mean_price, color = ~genre, 
    type = 'scatter', mode = 'lines+markers',
    name = ~paste("Mean Price -", genre),
    line = list(dash = 'solid')
  ) %>%
  add_trace(
    x = ~year, y = ~median_price, color = ~genre, 
    type = 'scatter', mode = 'lines+markers',
    name = ~paste("Median Price -", genre),
    line = list(dash = 'dash')
  ) %>%
  layout(
    title = "Sales Price Over Time by Genre (Mean and Median)",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Sales Price (USD)"),
    hovermode = "closest",
    legend = list(orientation = "h", y = -0.3)
  )

plot
```


These will be graded similar to last time:

Is the type of graph & choice of variables appropriate to answer the question?
Is the graph clear and easy to interpret?
Is the graph publication ready?

## Problem 3 - data.table
Repeat Problem 1 from PS04 using data.table.

```{r}
library(data.table)
library(nycflights13)
```


### a. Generate a table (which can just be a nicely printed tibble) reporting the mean and median departure delay per airport. Generate a second table (which again can be a nicely printed tibble) reporting the mean and median arrival delay per airport. Exclude any destination with under 10 flights. Do this exclusion through code, not manually.

Additionally,

Order both tables in descending mean delay.
Both tables should use the airport names not the airport codes.
Both tables should print all rows.

```{r}
# Departures delay
flights <- data.table(flights)
setDT(flights)
flights[, faa := dest]
```

```{r}
merged <- merge(flights[, faa := origin],
                airports,
                by = "faa",
                all.x = TRUE)
```

```{r}
merged[, .(N = .N,
           mean_delay = mean(dep_delay, na.rm = TRUE),
           median_delay = median(dep_delay, na.rm = TRUE)),
       by = name] |>
  _[N >= 10, !"N"] |>
  _[order(mean_delay, decreasing = TRUE)]
```

```{r}
# Departures delay
flights <- data.table(flights)
setDT(flights)
setDT(airports)
flights[, faa := dest]

# Arrival delay
merged <- merge(flights[, faa := dest],
                airports,
                by = "faa",
                all.x = TRUE)
```

```{r}
merged[, .(name = ifelse(is.na(first(name)), first(faa), first(name)),
           N = .N,
           mean_delay = mean(arr_delay, na.rm = TRUE),
           median_delay = median(arr_delay, na.rm = TRUE)),
       by = faa] |>
  _[N >= 10, !c("faa", "N")] |>
  _[order(mean_delay, decreasing = TRUE)] |>
  print(x = _, nrows = 10000)
```



### b. How many flights did the aircraft model with the fastest average speed take? Produce a tibble with 1 row, and entires for the model, average speed (in MPH) and number of flights.

```{r}
planes <- data.table(planes)
setDT(planes)
```


```{r}
merged <- merge(flights,
                planes,
                by = "tailnum",
                all.x = TRUE)
```

```{r}
allmodels <- merged[, `:=`(num_flights = .N,
                           avg_mph = mean(distance/(air_time/60), na.rm = TRUE)),
                    by = model]

allmodels[allmodels[, .I[which.max(avg_mph)]],.(model, avg_mph, num_flights)]
```






